The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-01 kalman_walk_paper.tex
general edits to paper. Changed figure 6 (f2 vs f1 divergence)

\begin{abstract}
It is known that even if a species' phenotype has remained unchanged over evolutionary time,
the underlying mechanism may have changed,
since distinct molecular pathways can realize identical phenotypes.
In this paper, we use quantitative genetics and linear systems theory to study
how a gene network underlying a conserved phenotype
evolves as genetic drift of small mutational tweaks to the molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
In this setting we treat organisms as ``black boxes'' for which the environment provides input
and the phenotype is the output,
and there exists an exact characterization of the set of all mechanisms that give the same input--output relationship. 
We show that in this situation, there is never a unique architecture for any phenotype
and that the evolutionary exploration of these distinct and mutationally connected mechanisms
can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations.
We estimate that in reasonably numerous species,
this process could thus the formation of new species
over a number of generations proportional to the effective population size.
This model also provides a natural explanation for Haldane's rule
that appears mechanistically distinct from other proposed mechanisms.
\end{abstract}

\plr{Additional ideas to consider adding:}
\begin{itemize}
    \item speciation literature
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    %\item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
    \item ``On the origin of species not by means of natural selection''
\end{itemize}

@@

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
these concepts can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
  \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype -- that is the population evolves under constant selective and environmental pressures.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
%Even when stabilizing selection preserves phenotype over evolutionary time, the underlying mechanism may change, 
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
%where the optimal phenotype remains constant over evolutionary time,
%so that despite strong stabilizing selection for phenotype,
%neutral drift in the underlying genotype is possible.
Applying results from system theory, we present
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes, and show that any biological system, in principal, can undergo system drift, as there is never a unique system architecture per phenotype.
%which gives concrete expectations for how any such
%system can, in principal, undergo system drift and rewire.
Even under stabilizing selection, a population can explore the set of all possible phenotypically equivalent gene networks.
Phenotypically equivalent gene networks are not necessarily compatible with one another, thus system drift can rapidly lead to the accumulation of hybrid incompatibilities analogous to those in the Dobzhansky-Muller model. 
Consequentially, two populations isolated for a sufficiently long period of time, 
will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

\section*{Methods}

Here we construct as abstract model to study biological systems, such as gene regulatory networks. The behavior (i.e. temporal dynamics) of such a system is determined both by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
%We study an abstract model of temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
%within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The $m$ ``inputs'' determined exogenously to the system are denoted $u(t)$,
and the $\ell$ ``outputs'' are denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_i(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$;
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations of the other concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ an $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions,
we can write the phenotype as
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

@@

% {Linearity} also ``zero-state equivalence''
Here we have assumed that the system is linear,
and begins from the ``zero'' state ($\kappa(0)=0$).
Of course, neither of these are necessarily true for real systems,
but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
Furthermore, the ease of analyzing linear systems makes this an attractive place to start.

@@

      Suppose the input is an impulse at time zero (a delta function), and so its phenotype is equal to its impulse response,
      \begin{align*}
        \phi(t) = h(t) = \sin t + \cos t  .
      \end{align*}
      The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
    \begin{figure}[H]
       % \begin{center} 
      \centering
         \begin{tabular}{cc}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (4,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (2,2) {input ($u(t)$)};
                \node[shape=rectangle] (y) at (2,-2) {output ($\phi(t)$)};
            \end{scope}

            \begin{scope}[>={Stealth[black]},
                          every node/.style={fill=white,circle},
                          every edge/.style={draw=black, thick}]
                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
            \end{scope}
            \begin{scope}[>={Stealth[black]},
                          every edge/.style={draw=black, thick}]
                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
            \end{scope}

            \end{tikzpicture} &
      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
   \end{tabular}
  %  \end{center}
      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.} \label{fig:oscillator}
    \end{figure}
  \end{example}

@@

Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, 
and also still preserve their phenotypes.
More generally, even if the system is not minimal, 
results from systems theory % -- the Kalman decomposition --
explicitly describe the set of all phenotypically equivalent systems.
% Could define script S here for 'system' if we want: $\Sys$
We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
to the system defined by $(A_0, B_0, C_0)$.
Concretely, this is
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \allS(A_0, B_0, C_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  .
  \end{aligned}
\end{equation}
These systems need not have the same kryptotypic dimension $n$,
but must have the same input and output dimensions ($\ell$ and $m$, respectively).

@@

Finally, note that if $B$ and $C$ are held constant --
i.e., if the relationships between environment, kryptotype, and phenotype do not change --
there are \emph{still} usually degrees of freedom. 
These correspond to distinct genetic networks that perform indistinguishable functions. 
The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of example \ref{ex:oscillator},
that all share common $B$ and $C$ matrices.

@@

    \begin{figure}[H]
    \centering
\begin{tabular}{cc}
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input $(u)$};
    \node[shape=rectangle] (y) at (2,-2) {output $(\phi)$};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
    \path[->] (U) edge node {\tiny $1$} (A);
    \path[->] (U) edge node {\tiny $1$} (B);
    \path[->] (A) edge[bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
\end{scope}
\end{tikzpicture} &	\includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
    \end{tabular}
% \caption{}
      \caption{
      (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
      % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
      } 
    \label{fig:all_osc}
\end{figure}
%
\begin{figure}[H]
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node (B) at (0,4) {$\kappa_{2}$};
    \node (C) at (3.464,2) {$\kappa_{3}$};
    \node[shape=rectangle, rotate=0] (U) at (-2.5,2) {input $(u)$};
    \node[shape=rectangle] (y) at (3.464,0) {output $(\phi)$};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny ?} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
    \path [->, sloped] (A) edge[bend right] node {\tiny ?} (C); 
    \path [->, sloped] (B) edge[bend left] node {\tiny ?} (C); 
    \path [->, sloped] (C) edge[bend right] node {\tiny ?} (A); 
    \path [->, sloped] (C) edge[bend left] node {\tiny ?} (B); 
    \path[->] (U) edge node {\tiny $1$} (A);
    \path[->] (U) edge node {\tiny $1$} (B);
    \path[->] (A) edge node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop below] node[sloped, anchor=center, below] {\tiny ?} (A);
    \path [->] (B) edge[loop above] node[sloped, anchor=center, above] {\tiny ?} (B);
    \path [->] (C) edge[loop right] node[sloped, anchor=center, above] {\tiny ?} (C);
\end{scope}
\end{tikzpicture}
  \caption{A possible non-minimal three-gene oscillator phenotypically equivalent to the systems in Examples \ref{ex:oscillator} and \ref{ex:all_osc}.}
  \label{fig:3D_osc}
\end{figure}

\paragraph{Sexual reproduction and recombination}
Parents with phenotypically equivalent, yet different, gene networks may produce offspring with dramatically different phenotypes. 
If the phenotypes are significantly divergent then the offspring may be inviable or otherwise have very low fitnesses, 
despite both parents being well adapted. If this is consistent for the entire population, we would consider them to be separate species.

Diploid organisms have two copies of the genome, each of which encodes a set of system coefficients.
We assume that a diploid who has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from her mother and father, respectively,
has phenotype determined by the system that averages these two,
$((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.

Each of the copies of the genome an organism inherits from her parents are generated from their two copies of the genome by meiosis,
in which the diploid parent recombines her two genomes.
We will assume that each coefficient (i.e., entry of $A$, $B$ or $C$)
is determined by a single nonrecombining locus,
so that each coefficient in the system produced by meiosis is an independent random choice
of the two parental coefficients.
With these definitions, an $F_1$ offspring of two individuals
carries one system copy from each parent,
and an $F_2$ is the offspring of two independently formed $F_1$ individuals.
If the parents are from distinct populations,
these are first- and second-generation hybrids, respectively.

@@

\section*{Examples}
\begin{example}[Metabolic network] \label{ex:metabolic}

>>>>

\begin{abstract}
%It is known that even if a species' phenotype has remained unchanged over evolutionary time,
%the underlying mechanism may have changed,
%since distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
Here we use quantitative genetics and linear system theory to study
how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small mutational tweaks to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
In this setting we treat an organism as a ``black box''
for which the environment provides input and the phenotype is the output,
%which translates some environmental input
%to an output that is subject to the pressures of natural selection-- the phenotype.
and there exists an exact characterization of the set of all mechanisms that give the same input--output relationship.
%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
Within this framework, we show that there is never a unique network architecture for any phenotype
and that the evolutionary exploration of these distinct and mutationally connected mechanisms
can lead to the reproductive incompatibility of idependently evolving populations.
  This evolutionary exploration, which we refer to as system drift, proceeds at a rate proportional to the effective population size ($N_e$) and amount of intrapopulation genetic variation.
We estimate that %in reasonably numerous species,
  this process can lead to the formation of new species, perhaps in as few as $0.1 N_e$ generations.
%over a number of generations proportional to the effective population size.
This model also naturally yields, what appears to be, a distinct explanation for Haldane's rule, or why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
\end{abstract}

%\plr{Additional ideas to consider adding:}
%\begin{itemize}
%    \item speciation literature
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    %\item hybrid vigor (need to do calculation)
 %   \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
 %   \item ``On the origin of species not by means of natural selection''
%\end{itemize}

@@

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
 % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype -- that is the population evolves under constant selective and environmental pressures.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
%Even when stabilizing selection preserves phenotype over evolutionary time, the underlying mechanism may change, 
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
%where the optimal phenotype remains constant over evolutionary time,
%so that despite strong stabilizing selection for phenotype,
%neutral drift in the underlying genotype is possible.
Applying results from system theory, we present
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes, and show that any biological system, in principal, can undergo system drift, as there is never a unique system architecture per phenotype.
%which gives concrete expectations for how any such
%system can, in principal, undergo system drift and rewire.
Even under stabilizing selection, a population can explore the set of all possible phenotypically equivalent gene networks.
Phenotypically equivalent gene networks are not necessarily compatible with one another, as such system drift may result in the reproductive incompatibility between populations isolated for a sufficiently long period of time, even in the absence of any sort of adaptative, selective, or environmental change. Finally, in some cases we estimate that reproductive incompatibility due to system drift can manifest on timescales on the order of $N_e$ generations.

analogous to those in the Dobzhansky-Muller model. 
%Consequentially, two populations isolated for a sufficiently long period of time, 
%will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

\section*{Evolutionary system theory}

Here we construct as abstract model to study biological systems, such as gene regulatory networks. The behavior (i.e. temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
%We study an abstract model of temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
%within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The $m$ ``inputs'' determined exogenously to the system are denoted $u(t)$,
and the $\ell$ ``outputs'' are denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_i(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$;
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations of the other concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ an $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions,
we can write the phenotype as
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

@@

% {Linearity} also ``zero-state equivalence''
Here we have assumed that the system is linear,
and begins from the ``zero'' state ($\kappa(0)=0$).
Of course, neither of these are necessarily true for real systems,
but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
To demonstrate this approach, we apply it to construct a simple gene network in Example \ref{ex:oscillator} below.

@@

      Suppose the input is an impulse at time zero (a delta function), and so its phenotype is equal to its impulse response,
      \begin{align*}
        \phi(t) = h(t) = \sin t + \cos t  .
      \end{align*}
      The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
    \begin{figure}[H]
       % \begin{center} 
      \centering
       %  \begin{tabular}{cc}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node (B) at (4,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (2,2) {input ($u$)};
                \node[shape=rectangle] (y) at (2,-2) {\begin{tabular}{cc} output ($\phi$) \\  \includegraphics[width=0.15\textwidth, height=0.04\paperheight]{osc_impulse} \end{tabular}};
            \end{scope}

            \begin{scope}[>={Stealth[black]},
                          every node/.style={fill=white,circle},
                          every edge/.style={draw=black, thick}]
                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
                \path[->] (U) edge node {\tiny $1$} (A);
                \path[->] (U) edge node {\tiny $1$} (B);
                \path[->] (A) edge[bend right] node {\tiny $1$} (y);
            \end{scope}
            \begin{scope}[>={Stealth[black]},
                          every edge/.style={draw=black, thick}]
                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
            \end{scope}

            \end{tikzpicture} %&
   %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
   %\end{tabular}
  %  \end{center}
      \caption{
        %(Left) 
        Diagram of the gene network in Example \ref{ex:oscillator}
      %, and (right) plot of the phenotype $\phi(t)$ against time $t$
      .} \label{fig:oscillator}
    \end{figure}
  \end{example}

@@

Some gene networks, however, can grow or shrink, perhaps following gene duplications and deletions, 
and also still preserve their phenotypes.
More generally, even if the system is not minimal, 
results from systems theory % -- the Kalman decomposition --
explicitly describe the set of all phenotypically equivalent systems.
% Could define script S here for 'system' if we want: $\Sys$
We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
to the system defined by $(A_0, B_0, C_0)$.
Concretely, this is
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \allS(A_0, B_0, C_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  .
  \end{aligned}
\end{equation}
These systems need not have the same kryptotypic dimension $n$,
but must have the same input and output dimensions ($\ell$ and $m$, respectively).

@@

Finally, note that if $B$ and $C$ are held constant --
i.e., if the relationships between environment, kryptotype, and phenotype do not change --
there are \emph{still} usually degrees of freedom. 
These correspond to distinct genetic networks that perform indistinguishable functions. 
The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of Example \ref{ex:oscillator},
that all share common $B$ and $C$ matrices.
The oscillator can also be equivalently realized by a three-gene (or larger) network, and will have even more evolutionary degrees of freedom available, as in e.g., Figure \ref{fig:3D_osc}.

@@

    \begin{figure}[H]
    \centering
\begin{tabular}{cc}
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input ($u$)};
    \node[shape=rectangle] (y) at (2,-2) {output ($\phi$)};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
    \path[->] (U) edge node {\tiny $1$} (A);
    \path[->] (U) edge node {\tiny $1$} (B);
    \path[->] (A) edge[bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
\end{scope}
\end{tikzpicture} &	\includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
    \end{tabular}
% \caption{}
      \caption{
      (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
      % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
      } 
    \label{fig:all_osc}
\end{figure}
%
\begin{figure}[H]
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node (B) at (0,4) {$\kappa_{2}$};
    \node (C) at (3.464,2) {$\kappa_{3}$};
    \node[shape=rectangle, rotate=0] (U) at (-2.5,2) {input ($u$)};
    \node[shape=rectangle] (y) at (3.464,0) {output ($\phi$)};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2.2$} (B);
   % \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
    \path [->, sloped] (A) edge[bend right] node {\tiny $2$} (C); 
    %\path [->, sloped] (B) edge[bend right] node {\tiny ?} (C); 
    \path [->, >=Rectangle, sloped] (C) edge[bend right] node {\tiny $-1$} (A); 
    \path [->, >=Rectangle, sloped] (C) edge[bend right] node {\tiny $-2.2$} (B); 
    \path[->] (U) edge node {\tiny $1$} (A);
    \path[->] (U) edge node {\tiny $1$} (B);
    \path[->] (A) edge node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop below] node[sloped, anchor=center, below] {\tiny $1$} (A);
    \path [->] (B) edge[loop above] node[sloped, anchor=center, above] {\tiny $1$} (B);
    \path [->, >=Rectangle] (C) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1$} (C);
\end{scope}
\end{tikzpicture}
  \caption{A possible non-minimal three-gene oscillator phenotypically equivalent to the systems in Examples \ref{ex:oscillator} and \ref{ex:all_osc}.}
  \label{fig:3D_osc}
\end{figure}

\paragraph{Sexual reproduction and recombination}
Parents with phenotypically equivalent yet differently wired gene networks may produce offspring with dramatically different phenotypes. 
If the phenotypes are significantly divergent then the offspring may be inviable or otherwise dysfunctional, 
despite both parents being well adapted. If this is consistent for the entire population, we would consider them to be separate species, in accord with the biological species concept \citep{mayr2000biological}.

Diploid organisms have two genomes, each of which encodes a set of system coefficients.
We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from each of its parents,
has phenotype determined by the system that averages these two,
$((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.

Each genome an organism inherits is generated by meiosis,
in which both of its diploid parents recombine their two genomes.
We will assume that each coefficient (i.e., entry of $A$, $B$ or $C$)
is determined by a single nonrecombining locus,
so that each coefficient in the system produced by meiosis is an independent random choice
between the two parental coefficients.
With these definitions, an $F_1$ offspring
carries a system copy from each parent,
and an $F_2$ is an offspring of two independently formed $F_1$s.
If the parents are from distinct populations,
these are simply first-- and second--generation hybrids, respectively.

@@

%\section*{Examples}
%\begin{example}[Metabolic network] \label{ex:metabolic}
%Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
